---
output:
  BiocStyle::html_document
---

<!---
Because we split the analysis pipeline in different independent files,
to speed up processing it, here in the setup block we load libraries and
objects that were loaded or produced in the previously processed file,
and which are necessary in this file.
--->



```{r setup, cache=FALSE, echo=FALSE, results='hide', message=FALSE}
library(knitr)
library(SummarizedExperiment)
library(edgeR)
library(geneplotter)
library(limma)
library(calibrate)
library(ggplot2)

opts_chunk$set(cache=TRUE,
               cache.path="cache/file2",
               cache.extra=R.version.string,
               autodep=TRUE,
               fig.align="center",
               comment="")

coadse.tumor <- readRDS(file.path("results", "coadse.tumor.filt.rds"))
dge.tumor <- readRDS(file.path("results", "dge.tumor.filt.rds"))
```

# Differential expression

After having filtered and normalized our tumor dataset, we looked for differential expression of genes among the cancer stages.
We applied a limma voom pipeline, the steps of which are explained in depth already in the paired DEanalysis section and for this reason here we only focus on the steps that differ in the new one.
Since the tumor stage has 4 different values ("StageI","StageII","StageIII", "StageIV""), we decide to use a model with no intercept and a contrast matrix instead, following the instruction of the "Several groups" section of the [limma user guide](http://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf).
After having obtained the model with the help of the contrast matrix, we applied eBayes. After multiple test adjusting, we are left with no DE genes. 
```{r}
coadse.tumor$ajcc_pathologic_tumor_stage <- gsub(x = coadse.tumor$ajcc_pathologic_tumor_stage, pattern = " ", replacement = "")
coadse.tumor$ajcc_pathologic_tumor_stage <- droplevels(as.factor(coadse.tumor$ajcc_pathologic_tumor_stage))
coadse.tumor$age_at_initial_pathologic_diagnosis <- droplevels(coadse.tumor$age_at_initial_pathologic_diagnosis)
lev <- c("StageI","StageII","StageIII", "StageIV")
f <- factor(coadse.tumor$ajcc_pathologic_tumor_stage, levels=lev)
design <- model.matrix(~0+f , colData(coadse.tumor))
colnames(design) <- lev
fit1 <- lmFit(assays(coadse.tumor)$logCPM,design)
contrast.matrix <- makeContrasts(StageI-StageII, StageI-StageIII, StageI-StageIV, StageII-StageIII, StageII-StageIV, StageIII-StageIV, levels=design)
fit2 <- contrasts.fit(fit1, contrast.matrix)
fit2 <- eBayes(fit2)
tt<-topTable(fit2, coef=1, n=Inf,adjust="fdr")
DEgenes <- rownames(tt)[tt$adj.P.Val < 0.1]
length(DEgenes)
saveRDS(tt, file.path("results", "tt.rds"))
```

This can be due to the fact that in many different biological contexts gene expression changes may be very little and after Multiple test adjusting no significant DE genes can be identified.
We then decide to run a Principal Component Analysis to further explore the gene expression among the four cancer stages.
Since we use the prcomp() function, we first need to transpose our matrix in order to have the samples as rows, if not we would get a pca analysis explaining how genes are related to each other, which in this case is not the desired output. After having prepared our data, we run the pca analyis and we can observe that the first four principal components explain respectively 11.7%, 7%, 6% and 4% of the variability. 
```{r pca, fig.cap="PC 1 and PC 2 by cancer Stage"}
a<-assays(coadse.tumor)
ta<-t(assays(coadse.tumor)$logCPM)
pca<- prcomp(ta)
summary(pca)
dtp <- data.frame('Stages' = coadse.tumor$ajcc_pathologic_tumor_stage, pca$x[,1:2]) # the first two componets are selected (NB: you can also select 3 for 3D plottings or 3+)
ggplot(data = dtp) + geom_point(aes(x = PC1, y = PC2, col = Stages)) + theme_minimal() 
```
In Figure @ref(fig:pca) it seems that there is no differential expression among the stages. However, we decide to proceed with the GSE Analysis. In fact, there may be small but consistent changes in gene expression values which could still imply relevant changes between the different stages.

## Session information

```{r, message=FALSE}
sessionInfo()
```
