---
output:
  BiocStyle::html_document
---

<!---
Because we split the analysis pipeline in different independent files,
to speed up processing it, here in the setup block we load libraries and
objects that were loaded or produced in the previously processed file,
and which are necessary in this file.
--->



```{r setup, cache=FALSE, echo=FALSE, results='hide', message=FALSE}
library(knitr)
library(SummarizedExperiment)
library(edgeR)
library(geneplotter)
library(sva)
library(limma)
library(calibrate)
library(EnhancedVolcano)

opts_chunk$set(cache=TRUE,
               cache.path="cache/file2",
               cache.extra=R.version.string,
               autodep=TRUE,
               fig.align="center",
               comment="")

coadse.filt <- readRDS(file.path("results", "coadse.filt.rds"))
dge.filt <- readRDS(file.path("results", "dge.filt.rds"))
```

# Differential expression

After the data filtering and normalization, we are interested in identifying gene expression changes and associated p-values between tumor and control samples. For this task, we used the R/Bioconductor packages [sva](http://bioconductor.org/packages/sva) and [limma](https://bioconductor.org/packages/release/bioc/html/limma.html).
We are adopting linear regression models, which in general are used in order to represent, in the most accurate way possible, something which is very complex. In the specific case of DE analysis, the model is used as a predictor of gene expression values. 
For the purpose of fitting the said regression model, it is crucial to build a design matrix. The design matrix is a n x m matrix, where n is the number of samples and m is the number of coeffients which need to be estimated or the types of samples, like cases and controls.
It needs to be defined for which factors to adjust. In our case we are working with a paired design and we would have liked to adjust for gender.

We decide to proceed by only adjusting for the patient id, as it is already implicitly adjusting for gender; as we are working with a paired design, gender is identical between every pair of samples, which would make reduntand to adjust for it and would create technical problems during the inversion step of the matrix in the least square algorithm steps.

In the follwing step we create the design matrix, only adjusting for the patient id. We verify as well wether the matrix is full rank and so is invertible with no problem, which is the case.
```{r}
pid <- substr(colnames(dge.filt), 9, 12)
mod <- model.matrix(~ factor(type) + factor(pid), colData(coadse.filt))
mod0 <- model.matrix(~ factor(pid), colData(coadse.filt))
#Verify wether is full rank --> yes!
is.fullrank(mod)
```
We use the [sva package](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3307112/) for identifying and removing batch effects and other unwanted sources of variation.
```{r}
sv <- sva(assays(coadse.filt)$logCPM, mod, mod0)
sv$n
```

The SVA algorithm has found `r sv$n` surrogate variables which we add to our model.

```{r}
modsv <- cbind(mod, sv$sv)
mod0sv <- cbind(mod0, sv$sv)
```
In order to compare gene expression values between samples we need to adjust for the mean-variance realtionship.

```{r libsize, fig.cap="Library sizes among samples"}
barplot(sort(dge.filt$samples$lib.size)/1e+06, xlab = "Samples", ylab = "Library sizes (Millions)", col=rgb(0.7, 0.1, 0.3))
```

This step could be executed with both limma trend and limma voom. We decided to proceed with the voom() function, since as we can observe in the Figure \@ref(fig:libsize), there are big differences in library sizes among the samples.

So, we calculate the weights that estimate the mean-variance realtionship at gene-by-sample level with the `voom()` function. The obtained weights will be used to fit the model. 

```{r, fig.cap="Mean-variance trend"}
v <- voom(dge.filt, modsv, plot=TRUE)
```

Next, we fit a linear model with the `fit()` function.

```{r}
# Fit linear model for each gene given a series of arrays
fit<- lmFit(v,modsv)
```

We then calculate the moderated t-statistics through the `eBayes()` function.

```{r}
# compute moderated t-statistics, moderated F-statistic, and log-odds of differential expression by empirical Bayes moderation of the standard errors towards a common value.
fit <- eBayes(fit)
FDRcutoff <- 0.01
```

Next, we classified the obtained test statistics as up, down or not significant with the help of the `decideTest` function. The FDR threshold used for this test is corresponds to 1%, as in cancer data is usual to have a lot of changes and so we prefer to be strict.

```{r}
#Classify a series of related t-statistics as up, down or not significant. 
res <- decideTests(fit, p.value = FDRcutoff) 
summary(res)
```

We extract the genes chromosome to look at the distribution of the differential expressed genes among the chromosomes, again for the purpose of getting an overview of the data we will be working with.

```{r chrs, fig.cap="Distribution of DE genes among chromosomes"}
genesmd <- data.frame(chr = as.character(sub("\\_.*", "", seqnames(rowRanges(coadse.filt)))), symbol = rowData(coadse.filt)[, 1], stringsAsFactors = FALSE)
#Extract a table of the top-ranked genes from a linear model fit.
fit$genes <- genesmd
tt <- topTable(fit, coef = 2, n = Inf)
genes<-table(as.character(sub("\\chr*", "",tt$chr[tt$adj.P.Val < FDRcutoff])))
par(las=2) # make label text perpendicular to axis
par(mar=c(5,8,4,2)) # increase y-axis margin.
genesmd
names.arg=c("1","2","3","4","5","6","7","8","9","10","11","12","13","14","15","16","17","18","19","20","21","22","X","Y")
barplot(genes,  cex.names=0.7, names.arg = names.arg, col=rgb(0.7, 0.1, 0.3))
sort(table(tt$chr[tt$adj.P.Val < FDRcutoff]), decreasing = TRUE)
```

Now, we produce two diagnostic plots for limma DE analysis.

```{r dp, fig.cap="Diagnostic plots for limma DE analysis"}
par(mfrow = c(1, 2), mar = c(4, 5, 2, 2))
hist(tt$P.Value, xlab = "Raw P-values", main = "A)", las = 1, col=rgb(0.7, 0.1, 0.3))
qqt(fit$t[, 2], df = fit$df.prior + fit$df.residual, main = "B)", pch = ".", cex = 3) 
abline(0, 1, lwd = 2, col= "red")
```

In the A plot of figure \@ref(fig:dp) we visualize the raw p-values distribution: it is mainly uniform a part from the very high peak on the left which corresponds to the significantly DE genes. In the B Q-Q plot of figure \@ref(fig:dp) we observe that our gene expression values are not following a normal distribution. In fact, if that was the case, we would expect the two lines overlapping and no DE genes. 

One interesting aspect we would like to systematically assess is the accuracy of the model with different combinations of adjusting factors and when using different algorithms when adjusting for the mean variance relationship. 
For doing this, we would need to use a dataset of genes which are known to be differentially expressed between patients with colon cancer and  control individuals with no colon cancer.
At the moment, we could not find a suitable dataset for this task but in the future, to expand the analysis, it would be interesting to investigate this aspect further.
Lastly, we just collect our set of differential expressed genes to proceed with further analyses.

We proceed to filter only those genes with FDR < 1% and a logFC > 2 (a minimum 100% change in expression)

```{r}
tt.filt <- tt[tt$adj.P.Val < FDRcutoff,]  #filtering by FDR < 0.01
tt.filt <- tt.filt[abs(tt.filt$logFC)>2,] # filtering by logFC > 2
all_DEgenes <- tt.filt
up_regulated <- all_DEgenes[which(all_DEgenes$logFC > 0),] # DE genes up regulated
down_regulated <- all_DEgenes[which(all_DEgenes$logFC < 0),] # DE genes down regulated
length(rownames(all_DEgenes)) # how many DE genes we have
length(rownames(up_regulated)) #  Up-reg DE
length(rownames(down_regulated)) # Down-reg DE

#down regultated in normal - up regulated in tumor
head(down_regulated, n = 30 )

head(up_regulated, n = 30 )
#Looking at individual genes, upregulation of connective tissue genes was prevalent (e.g.,??COL6A3, COL1A2, COL12A1, COL5A2, #COL3A1, CTHRC1, SULF1), as were genes involved in extracellular matrix function (e.g.,??LAMA4, PI3, POSTN, TIMP2) and cell #adhesion (e.g.,??TNS1). Genes involved in endothelial function (e.g.,??TIMP2) and specifically, colon cancer tumour #endothelium, such as the anthrax toxin receptor (ANTXR1), were also upregulated (Liu??et al, 2008).

# upregulation of connective tissue genes
all_DEgenes$symbol[grep("^COL", all_DEgenes$symbol)]
up_regulated$symbol[grep("^COL", up_regulated$symbol)]
down_regulated$symbol[grep("^COL", down_regulated$symbol)]
#AnnotationDbi::select(org.Mm.eg.db, keys=genes, columns='ENTREZID', keytype='SYMBOL')
down_regulated <- down_regulated[order(down_regulated$logFC),]
up_regulated <- up_regulated[order(up_regulated$logFC, decreasing = TRUE),]
down_regulated
up_regulated
head(down_regulated$symbol, n = 8)
head(up_regulated$symbol, n = 8)
head(coadse.filt$type)
sort(down_regulated[down_regulated$logFC], decreasing = TRUE)
##genes <- c("LAMA4", "PI3","POSTN"," TIMP2")
"CTNNB1"  %in% all_DEgenes$symbol
c("AGTR2","TP53","PIK3CA","AKT1","PDPK1","SMAD4","CTNNB1", "RET-ELE1")  %in% all_DEgenes$symbol
#%in% all_DEgenes$symbol
#c("COL6A3", "COL1A2", "COL12A1", "COL5A2", "COL3A1", "CTHRC1", "SULF1") %in% all_DEgenes$symbol


# Age related methylation 
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5175245/
c("PPP1R16B" )  %in% all_DEgenes$symbol

# GLYCOPROTEIN - HYPERMETHILATION
# https://www.uniprot.org/uniprot/O75493
# https://link.springer.com/article/10.1007/s10151-018-1820-3
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3123690/
c("CA11" )  %in% all_DEgenes$symbol
c("LVRN")  %in% all_DEgenes$symbol

#https://www.spandidos-publications.com/ol/17/4/3657

# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2897904/ :HDAC2
c("SOD1")  %in% all_DEgenes$symbol

c("SEPT","d","SYNE1", "DKK3", "SDC2", "SFRP1", "SYNE1", "CEMIP", "SPATA18" )  %in% all_DEgenes$symbol

# -- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5998821/
a <- c("SLC13A2", "NR1H4", "SLC9A3", "NXPE1", "WFDC2", "HOXA13", "AK1", "CYP2C18", "ST3GAL4", "APOA4", "GCG", "MUC5B", "TRPM6", "PI3", "SLPI", "FOXA2", "HOXD13", "PYY", "SLC14A2", "SPINK5", "CA1", "SLC37A2", "CHST5", "GALNT5", "SLC28A2", "SLC20A1", "AHSG", "CDHR1", "ADRA2A", "SPON1", "CLDN8", "HKDC1", "B3GNT7", "LRRC43", "PRAC1", "HOXB13", "ST6GALNAC6", "NLRP4", "CYP11B1", "NPHS1", "CAPN13", "PITX2", "C7orf57", "CPA6", "CYP2C19", "CKB", "ANPEP", "B4GALNT2", "DPCR1", "DRD5", "PKHD1", "KRT15", "REG3A", "INSL5", "HOXC5", "MFSD4", "C2orf73", "ARSJ", "B3GALT5", "FAM3B", "C21orf88", "FFAR4", "GLDN", "LHFPL3", "TMEM72", "NWD1", "WNT7B", "CTD-2228K2.5", "PLA2G2A", "ESRRG", "SULT1C2", "HSD3B2", "NPY4R", "MUC12", "GBP7", "PRAC2", "L1TD1")
a %in% all_DEgenes$symbol
#WFDC2, AK1 
# Molecular transport, cell-to-cell signaling and interaction, nervous system development and function
# 


b <- c("ADK", "AK1", " APP", "BCKDK", "CAB39L", "CCDC144NL", "CKMT1A/CKMT1B", "CXXC1", "ENO3", "ESRRG", " FGF12", "GSPT2", "H2AFJ", "H2AFY", "HSP90AA1", "LAMC1", "MAST2", "MOGAT3", " NOL4L", " NSRP1", "PRR15", " R3HDML", " RPS6KA6", " SLC13A3", " SMG1", "SMG5", "SMTNL2", " SMYD3", "SRPK1", "STK26", "STRADA", "TRIM7", " TTI2", " TTPA", " UPF3A")
b %in% all_DEgenes$symbol



# Wnt10a is upregulated in colon cancer tissues and is associated with clinicopathological factors


```
After the filtering, the total number of DE genes is `r length(rownames(all_DEgenes))`, from which `r length(rownames(up_regulated))` are up regulated and `r length(rownames(down_regulated))` are down regulated.

```{r plotMA, fig.cap="MA plot of DE genes with logFC > 2 and FDR cutoff < 0.01."}
tt.filt <- tt[tt$adj.P.Val < FDRcutoff,]  #filtering by FDR < 0.01
all_DEgenes <- tt.filt[abs(tt.filt$logFC)>2,] # filtering by logFC > 2
DEgenes <- rownames(all_DEgenes) 
limma::plotMA(fit, coef = 2, status = rownames(fit$lods) %in% DEgenes, legend = FALSE,
main = "MA plot", hl.pch = 46, hl.cex = 2, bg.pch = 46, bg.cex = 2, las = 1)
legend("bottomright", c("DE genes", "Not DE genes"), fill = c("red", "black"), inset = 0.01)
saveRDS(all_DEgenes , file.path("results", "alldegenes.rds"))
```
In figure \@ref(fig:plotMA) we can see the distribution of the filtered DE genes. One reassuring thing here is that the dots are centered around the value zero which indicates that there are not expression-level dependent biases.

## Volcano plot 

Lastly, we want to visualize the DE genes in a Volcano plot. We set two different thresholds: the log fold change threshold and the adjusted p-value threshold.

```{r, fig.cap= "Volcano plot"}
EnhancedVolcano(tt,
    lab = tt$symbol,
    x = 'logFC',
    y = 'adj.P.Val',
    xlim = c(-8, 8),
    title = 'Volcano Plot',
    pCutoff = 10e-16,
    FCcutoff = 2,
    transcriptPointSize = 1.0,
    transcriptLabSize = 3.0)
```

## Session information

```{r, message=FALSE}
sessionInfo()
```
