---
title: "Analysis of a TCGA RNA-seq data set on Colon Adenocarcinoma"
author:
- name: Altair C. Hernandez
  affiliation:
  - &id Master Programme on Bioinformatics for Health Sciences, Universitat Pompeu Fabra,   Barcelona, Spain
  email: altair.chinchilla01@estudiant.upf.edu
- name: Aina Rill
  affiliation: *id
  email: aina.rill01@estudiant.upf.edu
- name: Luisa Santus
  affiliation: *id
  email: luisa.santus01@estudiant.upf.edu


date: "`r format(Sys.time(), '%B %e, %Y')`"
output: 
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
bibliography: bibliography.bib
---

# Introduction

Colorectal Cancer (CRC) is the second most common cancer in women (614,000 cases per year) 
and the third most common in men(746,000 cases per year). Currently it is the most common 
malignant cancer in the gatrointestinal tract, representing 13\% of all malignant tumors, 
and it is considered the main cause of death in gastrointestinal cancer @Siegel2015.
It can be arised from one or a combination of three differen mechanisms, namely chromosmomal 
inestability (CIN), CpG islands methylator phenotype (CIMP), and microsatellite inestability (MSI). 
Understanding the specific mechanisms of tumorigenesis and the underlaying genetic and epigenetic
traits is crutial in the disease phenotype comprehension. [The Cancer Genome Atlas](https://www.cancer.gov/about-nci/organization/ccg/research/structural-genomics/tcga) (TCGA) has comprehensively profiled this type of cancer in a patient cohort. Here we analyze the expression profiles of those
patients, accessible in the form of a raw RNA-seq counts produced by
@rahman2015alternative using a pipeline based on the R/Bioconductor
software package `r Biocpkg("Rsubread")`.



```{r echo=FALSE, results='asis'}
cat(readLines("QAanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("DEanalysis.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("FEanalysis.knit.md"), sep="\n")
```

# Colorectal cancer Stages' analysis

```{r echo=FALSE, results='asis'}
cat(readLines("QAanalysisTumor.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("DEanalysisTumor.knit.md"), sep="\n")
```

```{r echo=FALSE, results='asis'}
cat(readLines("FEanalysisTumor.knit.md"), sep="\n")
```

# References
